---
title: "How to Make a Player Level Team Ratings and Win Projections"
author: "Andrew Patton"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## What are Player Level Team Ratings?

Team ratings are a way to determine the strength of a team based on player level ratings, often statistical plusminus models (SPM). For a primer on those, check out [this link](https://github.com/anpatton/basic-nba-tutorials/blob/main/spm/how_to_make_spm_R.md). Doing team ratings on the player level means that you are not just adding up points scored per 100 and subtracting points against per 100, as that is team performance data. The player level team ratings are often used for projection, gambling, etc. You might have seen versions of these from PIPM (RIP), DARKO, 538, etc. Like most basketball analytics approaches, there are a variety of ways to accomplish the task, and this is but one simplified version. Let's get into it. 

## What are Win Projections?

Simply, win projections are estimates of how many games a team will win over ___ period of time. Can be season long projections, daily projections, etc. Our projections will be based on our player level team ratings that we will calculate first.

## Data

1) player_ratings_minutes.csv" - This is DARKO player ratings from day one of the 2018-19 season with projected minutes as well.
2) "schedule.csv" - This is the league schedule from 2018-19.

```{r}
library(tidyverse)
library(knitr)

ratings <- read_csv("data/player_ratings_minutes.csv") %>% 
  mutate(dpm = o_dpm + d_dpm) %>% 
  arrange(-dpm) %>% 
  mutate(team_name = ifelse(team_name == "Los Angeles Clippers", "LA Clippers", team_name))

kable(ratings[1:10, ] %>% 
          mutate(across(where(is.numeric), function(x) round(x, 1)))) 

```

You'll note that Russ has an extremely high DPM. It starts to decrease somewhat quickly following this point due to three years of strong on-off data.

## Step 1: Aggregate player ratings to the team level and calculate team ratings

This is a deceptively simple step that does literally does everything! We multiply a player's minutes projection by their O-DPM and D-DPM, and then by team, add up the offensive and defensive values as well as the total minutes projected. Then, we divide the team offensive and defensive values by the teams' total projected minutes and multiply by five (for five players on court). That's it! Add up the offensive and defensive ratings and you have your net team ratings.

```{r}

team_ratings <- ratings %>% 
  replace_na(list(minutes = 0)) %>% 
  mutate(o_value = minutes * o_dpm) %>% 
  mutate(d_value = minutes * d_dpm) %>%
  group_by(team_name) %>% 
  summarise(team_o_value = sum(o_value),
            team_d_value = sum(d_value),
            team_minutes = sum(minutes)) %>% 
  mutate(ortg = (team_o_value/team_minutes) * 5,
         drtg = (team_d_value/team_minutes) * 5,
         nrtg = ortg + drtg) %>% 
  mutate(across(where(is.numeric), function(x) round(x, 1))) %>% 
  ungroup() %>% 
  select(team_name, ortg, drtg, nrtg) %>% 
  arrange(-nrtg) 

kable(team_ratings)

```

## Step 2: Get What we Need to Project Wins

In order to figure out the number of wins a team should have, we first need the schedule. I have scraped that already for the 2018-19 season in the "schedule.csv" and conveniently formatted it for our purposes. 

```{r, message=FALSE}

schedule <- read_csv("data/schedule.csv") 

head(schedule)

```

You'll note that home and away teams are there, as home teams are more likely to win. The raw team ratings assume a neutral court scenario. Let's join in the team ratings to the schedule.

```{r}

schedule_with_ratings <- schedule %>% 
  left_join(team_ratings, by = c("home_team" = "team_name")) %>% 
  left_join(team_ratings, by = c("away_team" = "team_name"), suffix = c("_home", "_away")) 

head(schedule_with_ratings)

```

Ok, with that done, our setup is ready to go. All we need to do is actually figure out how to translate ratings to win probability. We're going to cheat a bit here and use some previously conducted research on the topic by some of my NBA Twitter colleagues/friends/collaborators.  

The rating difference and win/loss was plugged into a logistic regression across many many games and home/away splits. The coefficients that resulted from that regression are used below as coef1 and coef2. The get_win_prob function takes the coefficients and manually conducts a prediction.

```{r}

get_win_prob <- function(home_rating, away_rating, sd) {
  
  hca <- 2.7 ## THIS IS DIFFERENT FOR COVID SEASONS
  home_rating_adj <- rnorm(mean = home_rating, sd = sd, n = 1)
  away_rating_adj <- rnorm(mean = away_rating, sd = sd, n = 1)
  delta <- home_rating_adj - away_rating_adj - hca
  coef1 <- 0.1483737284
  coef2 <- 0.4257284470
  home_win_prob <- 1/(1 + exp(-(coef2 + coef1 * delta)))
  
  return(home_win_prob)
  
}

```

You'll note that there is an "sd" parameter that we have not discussed. This is approximately the standard deviation of the team ratings based on the uncertainty in the player ratings based again on prior research. Essentially, when we say that Steph is a +7.0, we also imply that he is +7.0 +/- some value. No player is *exactly* what they are predicted at. The default here of 2.2 is also a value taken from previous research on the topic by not-me individuals.

Next, we need to build a function that simulates a season of games. This one is kind of hacky but does the trick. Like always, I'm sure there's a better/faster/simpler way to do some of this, but this is my first pass and it seems to work well. It takes in the schedule dataframe and the value for the sd and outputs a list of 1s and 0s where the home team is predicted to win across all the games. Important note - if we assume that the team ratings are normally distributed we can actually skip the simulation. However, I wanted to show the simulation code since that's probably more of use to people than various forms of _norm.

```{r}

simulate_season <- function(schedule, sd = 2.2) {
  
  home_rating_col <- which(names(schedule) == "nrtg_home")
  away_rating_col <- which(names(schedule) == "nrtg_away")
  home_win_probs <- unname(apply(schedule, 1, 
                                 function(x) get_win_prob(as.numeric(x[home_rating_col]),
                                                          as.numeric(x[away_rating_col]), 
                                                          sd = sd)))
  home_win_checks <- runif(n = length(home_win_probs))
  home_win <- as.numeric(home_win_probs >= home_win_checks)

  return(home_win)
  
}

```

Now, we're going to do a Monte Carlo (simulation) where we just run the simulation a few hundred times to see what shakes out. If we assumed the team ratings were deterministic, point predictions with no variace, this wouldn't be necessary. This is even more important when you aren't just doing Baby's First Win Projections.

We're going to do this in parallel as it's a perfect use case of completely separate trials that can be split across your cores. I'll annotate the code here if you're new to parallelization. 


```{r}

cores_minus_one <- parallel::detectCores() - 1 ## How many cores you have minus one (don't want to overdo it)
cl <- parallel::makeCluster(spec = cores_minus_one) ## Creating a cluster instance
parallel::clusterExport(cl, c("get_win_prob", "schedule_with_ratings", "simulate_season")) ## export our two functions and the data to the cluster
home_wins <- parallel::parSapply(cl, 1:100, function(i) simulate_season(schedule = schedule_with_ratings)) ## literally just sapply, but on our cluster for 100 trials
parallel::stopCluster(cl) ## close down the cluster when done

```

Great, that hopefully worked on your machine. Each column is a season and each row is a game. If you're a Mac user, check out parallel::mclapply which is specifically targeted for Monte Carlos but does not work on Windows due to CS terms over my head. Important note here, while we don't know the order in which the parallelized trials will occur, *within* the simulate season function, the game order is preserved.

Next, we tally up wins and losses by each team and do some summing

```{r}

home_wins_frame <- cbind(select(schedule_with_ratings, home_team), 
                         as.data.frame(home_wins)) %>% 
  group_by(home_team) %>% 
  summarise_all(sum) %>% 
  rename(team = home_team)


away_wins_frame <- cbind(select(schedule_with_ratings, away_team), 
                         as.data.frame(1 * !home_wins)) %>% 
  group_by(away_team) %>% 
  summarise_all(sum) %>% 
  rename(team = away_team)

wins_frame <- bind_rows(home_wins_frame, away_wins_frame) %>%
  ungroup() %>% 
  group_by(team) %>%
  summarise_all(sum)

```

With our one column per season and one team per row result, we can now take a look at the variation within a team how many games a they might win.

```{r}

projections <- apply(wins_frame[, -1], 1, function(x) data.frame(mean = mean(x),
                                                                 sd = sd(x))) %>% 
  bind_rows() %>% 
  bind_cols(wins_frame[, 1], .) %>% 
  mutate(best_case = qnorm(p = 0.95, mean = .$mean, sd = .$sd)) %>% 
  mutate(worst_case = qnorm(p = 0.05, mean = .$mean, sd = .$sd)) %>% 
  mutate(across(where(is.numeric), function(x) round(x, 1))) %>% 
  arrange(-mean)

kable(projections)  

```

And there you have it, win projections done as simply as possible!





